Resolvable Ambiguity

### 2019-07-10 paper/paper.tex
Formalism update

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
A common point of view when designing grammars for parsing is that
the grammar definition has to be unambiguous.  However, up front unambiguous grammars
can force language designers to make arbitrary choice to disambiguate
the language. In this paper, we depart from the traditional view of
unambiguous grammar design, and allow ambiguities to be delayed until
compile time, allowing the user to perform the disambiguation. A
natural decision problem follows: given a language definition, can a
user always disambiguate an ambiguous program?  We introduce and
formalize this fundamental problem---called the \emph{resolvable
  ambiguity problem}---and divide it into separate static and dynamic
resolvability problems. We prove soundness and completeness for a
restricted language in the static case, and soundness in the general
case for dynamic resolvability. The approach is evaluated through
two separate case studies, covering both a large existing
programming language, and the composability of domain-specific
languages.
\end{abstract}

@@

% Motivate why unambiguous parsing is a good (and important) idea.
Ever since the early 60s, the problem of deciding if a context-free
grammar is ambiguous has been known to be
undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
a large number of restricted grammars have been developed to guarantee
that a language definition is unambiguous. This traditional view of
only allowing unambiguous grammars has been taken for granted as the
only truth: \emph{the} way of how the syntax of a language must be
defined
%This seems to stem
%from the view that ambiguity has no place in programming language
%syntax, which is highly prevalent
~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
%
However, recent work by
~\citet{palmkvistCreatingDomainSpecificLanguages2019} shows why
carefully designed ambiguous syntax definitions are in some cases
preferable compared to completely unambiguous
definitions. Specifically, instead of forcing a language designers to
make arbitrary choice to disambiguate the language, a more natural
choice may be to postpone the decision to the programmer.

% Motivation II, compising DSLs
Another area---where ambiguous grammars naturally arises---is
domain-specific language development.  An important
objective of domain-specific language development is to construct
languages by composing and extending existing
languages.
Unfortunately, the composition of restricted
%grammars easily results in ambiguous grammars, which again are
%undecidable.
%
%constructing domain-specific programming languages there is often
%a desire to reuse work, to compose smaller language
%fragments.
%However, how to make such compositions well-behaved is not
%always obvious.
%In particular, when constructing new domain-specific languages from
%existing languages, by composing multiple syntax definitions,
grammars easily produces
an ambiguous grammar, even if all of the original grammars are
unambiguous. Approaches for solving this problem can be broadly split
into two categories: those that handle ambiguities on the
grammar-level (detection, prevention, etc.), and those that work on
particular examples of ambiguous programs. The former is well explored
in the form of heuristics for ambiguity detection
(e.g. \cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007})
or restrictions on the grammars to be composed
(e.g. \cite{kaminskiModularWellDefinednessAnalysis2013}), while the
latter has received relatively little attention.

%Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g. \cite{earleyEfficientCo%ntextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}. However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity. Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019} but not yet explored, is to automatically generate several unambiguous programs, one for each possible interpretation, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation. For example, in a language without precedence, \verb|'1 + 2 * 3'| is ambiguous, but the two possible interpretations can be written unambiguously as \verb|'(1 + 2) * 3'| and \verb|'1 + (2 * 3)'| respectively.

%In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists. For example, in a language with semicolon for both sequential composition and element separation in a list (e.g. OCaml), ``$[1; 2]$'' is ambiguous: it is either a list of two elements, or a list of one element, namely the sequential composition of $1$ and $2$. The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative.

% The problem

This paper focuses on the problems that arises if a grammar definition
is not guaranteed to be unambiguous. Specifically, if parsing a
program results in several parse trees, the grammar was obviously
ambiguous. The programmer then needs to disambiguate the program by,
for instance, inserting extra parenthesis. However, is it always
possible to resolve an ambiguous input program by augmenting the
program? We say that a program has a \emph{resolvable ambiguity} if
the ambiguity can be solved by augmenting the program. The \emph{resolvable ambiguity problem} can then be divided into several decision problems: (i) the dynamic decidability problem: given a program (or a

dynamic resolvable ambiguity problem of deciding if a context-free grammar is

This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar. We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot. We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}. This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.

@@

\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure-sound}
  A decision procedure $f$ solving the static resolvability problem is sound iff

@@

In practice, we may additionally be interested in a decision procedure
$g$ such that $g(P) \implies \lnot \rho(P) \text{ for all abstract
  parsers } P$. It turns out, we can give $g$ for the parse languages
studied in the following sections.

\begin{definition}[Completeness of Static Resolvability]\label{def:static-procedure-complete}
  A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff

@@

\noindent Similarly, we define soundness and completeness for the dynamic problem.

\begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure-sound}
  A decision procedure $f$ solving the dynamic resolvability problem is sound iff

  $f(\parse, w) \implies \rho_P(w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
\end{definition}

Analogous to the static case, for the parse languages studied in the
following sections we can also provide a decision procedure $g$ such
that $g(\parse, w) \implies \lnot \rho_P(w) \text{ for all } w \in
\Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$.

\begin{definition}[Completeness of Dynamic Resolvability]\label{def:dynamic-procedure-complete}
  A decision procedure $f$ solving the dynamic resolvability problem is complete iff

  $\rho_P(w) \implies f(\parse, w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
\end{definition}

@@

We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure-sound}, \ref{def:static-procedure-complete}, \ref{def:dynamic-procedure-sound}, and \ref{def:dynamic-procedure-complete}. Section~\ref{sec:static} gives a partial solution to the static problem, while Section~\ref{sec:dynamic} fully solves the dynamic problem with one caveat: we only consider languages with balanced parentheses.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
% Why?

Ever since the early 60s, the problem of deciding if a context-free
grammar is unambiguous or not has been known to be undecidable. As a
consequence, a large number of restricted grammars have been developed
to guarantee that a language definition is unambiguous. This
traditional view of only allowing unambiguous grammars has until
recently been taken for granted as the only truth: \emph{the} way of
how the syntax of a language must be defined. However, up front
unambiguous grammars can force language designers to make arbitrary
choice to disambiguate the language, although the natural choice may be to
postpone the decision to the programmer. Moreover, an important
objective of domain-specific language development is to construct
languages by composing and extending existing
languages. Unfortunately, the composition of restricted unambiguous
grammars easily results in ambiguous grammars, which again are
undecidable. In this paper, we depart from the traditional view of
unambiguous grammar design, and allow ambiguities to be delayed until
compile time, allowing the user to perform the disambiguation. A
natural decision problems follows: given a language definition, can a
user always disambiguate an ambiguous program?  We introduce and
formalize this fundamental problem---called the \emph{resolvable
  ambiguity problem}---and divide it into separate static and dynamic
resolvability problems. We prove soundness and completeness for a
restricted language in the static case, and soundness in the general
case for dynamic resolvability. The approach is evaluated through
three separate case studies, covering both a large
existing programming language, and the composability of
domain-specific languages.
%A traditional view when designing programming languages is that syntax
%definitions should be unambiguous: only unambiguous grammars are used
%to construct language specific parsers that either rejects a program,
%or accepts it and produces a unique parse tree. However,
\end{abstract}

@@

When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments. However, how to make such compositions well-behaved is not always obvious. In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous. Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous programs. The former is well explored in the form of heuristics for ambiguity detection (e.g. \cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g. \cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny. This seems to stem from the view that ambiguity has no place in programming language syntax, which is highly prevalent (e.g. \cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).

Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g. \cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}. However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity. Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019} but not yet explored, is to automatically generate several unambiguous programs, one for each possible interpretation, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation. For example, in a language without precedence, \verb|'1 + 2 * 3'| is ambiguous, but the two possible interpretations can be written unambiguously as \verb|'(1 + 2) * 3'| and \verb|'1 + (2 * 3)'| respectively.

In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists. For example, in a language with semicolon for both sequential composition and element separation in a list (e.g. OCaml), ``$[1; 2]$'' is ambiguous: it is either a list of two elements, or a list of one element, namely the sequential composition of $1$ and $2$. The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative. This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar. We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot. We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}. This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.

@@

% todo: improve label (where used?)
\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure}
  A decision procedure $f$ solving the static resolvability problem is sound iff

@@

In practice, we may additionally be interested in a decision procedure
$g$ such that $g(P) \implies \lnot \rho(P) \text{ for all abstract
  parsers } P$. It turns out, we can give $g$ for the parse languages
studied in following sections.

\begin{definition}[Completeness of Static Resolvability]
  A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff

@@

\noindent Similarly, for the dynamic problem:

\begin{definition}
  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the dynamic resolvability problem is:
  \begin{itemize}
  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$ and $w$.
  \end{itemize}
  \label{def:dynamic-procedure}
\end{definition}

@@

We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure} and \ref{def:dynamic-procedure}. Section~\ref{sec:static} gives a partial solution to the former problem, while Section~\ref{sec:dynamic} fully solves the latter with one caveat: we only consider languages with balanced parentheses.
